Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-05-30 manuscript/manuscript.Rmd
Making some coauthor edits before Anitra's review

This shift in community structure and productivity should result in a corresponding shift in the composition of the particulate organic matter produced. Eukaryotic organisms have distinct chemical fingerprints from those of cyanobacteria and heterotrophic bacteria [@Heal2021; @Durham2022] and the relative contribution of these different taxa to total biomass should reflect their chemical composition. However, recent work from @Harke2021 showed that overall community function may be robust across eddy types at the surface while @Gleich2024 detected significant differences in protistan community composition and metabolic potential down to 250 meters, with heterotrophy-associated transcripts found at higher abundances in the cyclone. These results raise questions about the effect of mesoscale features on particulate organic matter composition. [@Wei: You probably want to add several sentences here to briefly summarize what you did in this work to fill the gap. Often, you can propose a major hypothesis and mention the methods you used to analyze the samples to test your hypothesis.]

@@

Metabolomics samples from all three sample sets were extracted via a modified Bligh & Dyer [@Boysen2018], with the aqueous fraction run on a HILIC column paired with a Q-Exactive HF mass spectrometer. Metabolites for which an authentic standard was available and run alongside were integrated in Skyline (HEE and MESO-SCOPE eddy centers) and MSDIAL (MESO-SCOPE eddy transect). Untargeted analysis was performed using the R package XCMS with each feature manually inspected for quality assurance. [@Wei: I think this paragraph belongs to Methods section. @Will: Maybe just shorten it extensively? Ask Anitra for advice since Rachel wanted this here.]

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = .813, p-value < 0.001, Supplemental figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= .593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= .503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= .049$, p-value = 0.792). This meant that metabolites contributed a fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 2) [@Benedetto: I find this very informative. It would be great to get a ballpark sense of the percentages including the unknown metabolites, but I understand that it might be too uncertain considering the required assumptions on molecular formulas and concentration. I would like to know your wild guess, though]. Particulate nitrogen was tightly correlated with particulate carbon ($r=.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

@@

Given the strength of the metabolomic response to SLA at the DCM [@Benedetto: This seems to imply that you knew this before sampling], we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~40 meter depth range and the transition between the eddies explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001, Figure 4). Additionally, both sample depth and SLA were correlated with the first principal component of the metabolite matrix ($r_{depth}=.766$, $r_{SLA}=.684$, fraction of variance explained by PC1 = 30.8%). The cyclonic samples in particular show a much larger spread than the anticyclonic ones, indicating larger sample variability in the cyclone DCM relative to the anticyclone. A depth gradient is visible along NMDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom (Figure 4).

@@

fk_targ %>%
  filter(compound_name%in%c("Homarine", "Trigonelline", "Dimethylsulfonioacetate", "Dimethylsulfoniopropionate", "Taurine", "Isethionic acid", "Arsenobetaine")) %>%
  left_join(filled_file_metadata) %>%
  group_by(filename) %>%
  # mutate(nM=nM/sum(nM)) %>%
  ungroup() %>%
  filter(depth=="DCM") %>%
  # nest(data=-compound_name) %>%
  # mutate(aovout=map(data, ~broom::tidy(aov(.x$nM~.x$sla_class)))) %>%
  # unnest(aovout) %>% filter(term==".x$sla_class") %>%
  # mutate(p.value=p.adjust(p.value, method="fdr"))
  ggplot(aes(x=sla_class, y=nM)) +
  # geom_boxplot() +
  geom_point() +
  facet_wrap(~compound_name, scales="free_y")

@@

The eddy effects, on the other hand, were much less strong during this cruise. Of the six known metabolites with the strongest differences across the eddy transect at the DCM in the MESO-SCOPE cruise, only isethionic acid was significantly different in the HEE dataset (isethionic acid t-test p-value$_{FDR}$ = 0.043, other five p-values > 0.25) [@Wei: If P values are not significant, you can calculate effect size.]. This difference in isethionic acid concentration was only found after normalizing each sample to the sum of all metabolites in the sample to control for biomass. Arsenobetaine was again found to strongly enriched in the anticyclone at the DCM both when normalizing to biomass and when not, just as in MESO-SCOPE. Surprisingly, the 173.0921 *m/z* mass feature was also slightly enriched but this time in the *cyclone* DCM with peak areas approximately 1.4 times the values in the anticyclone (t-test p-value = 0.067, Mann-Whitney p-value = 0.045). The 170.1176 *m/z* mass feature was not detected in the HEE samples.

@@

The Hawaiian Eddy Experiment (HEE) data upset many of the expectations [@Wei: consider calculating the effect size, as mentioned above. @Will: Not sure what Wei means by this. Should I do effect size calculations here? Or just above when we didn't see an effect?] we developed during the MESO-SCOPE cruise in the previous year. Most surprising was the large difference between the eddies detected at the *surface*, while the DCM samples were functionally indistinguishable. This contrasts directly with the results from MESO-SCOPE and the results in @Gleich2024, who found eddy-driven shifts in protistan community composition to be larger at depth than at the surface. However, @Dugenne2023 found differences in nitrogen fixation rate and nitrogen fixer composition varying widely throughout the upper water column. The large surface differences were especially surprising given the large inter-replicate differences between the HEE surface samples in this study, while the DCM samples tended to be much more consistent.

@@

The strong SLA signal detected among the 175 meter samples was partially surprising to us, having expected the effect to be strongest at the DCM where eddy effects lift large concentrations of nutrients above the 1% light level. Instead, we found more compounds overall to be significantly different across the eddy transect in the 175 meter samples than we did at the DCM (α=0.05, 43 compounds at 175 meters vs 22 at the DCM) [@Benedetto: Notice that the analyses of SLA at Station ALOHA in Barone et al. (2019) clearly shows enhancement of organic matter and heterotrophic picoplankton below the DCM]. The enrichment of nucleobases in the 175 meter samples may hint at bacterial biomass, as nucleobases tend to be highly abundant in bacterial cultures [@Heal2021]. The abundance of a mass feature putatively identified as acetylcholine in the 175 meter cyclone samples followed the same general trend as arsenobetaine, both enriched at depth and in the anticyclone, and may be another compound that results from heterotrophic degradation of phytoplankton metabolites [@Durham2022]. These results raise important questions about the depth at which the community is isolated from SLA effects, with additional data from @Barone2022 and @Gleich2024 indicating that even at 250 meters eddy effects are still discernable.

@@

This study reports how changes in the biogeochemistry and community composition due to mesoscale eddies affect the metabolome and thereby the composition of particulate organic matter. A transect across adjacent mesoscale eddies of opposite polarity showed that many metabolites track closely with metrics of overall biomass, with eddy effects stronger at 175 meters than at 15 meters or the deep chlorophyll maximum (DCM). High-resolution depth sampling of the DCM at the center of each eddy elucidated several known and unknown biomarkers likely corresponding to the previously documented increase in eukaryotic phytoplankton in the cyclone. Metabolite clusters aligned well with expected trends in abundance with depth and between the two eddies, with rapid convergence below the DCM towards a general deep-sea metabolite signal [@Benedetto: with rapid convergence below the DCM towards a general deep-sea metabolite signal. @Will: Yeah, you right. I'll have to rephrase this.]. By contrasting these effects with a follow-up analysis in the centers of a separate pair of mesoscale eddies of opposite polarity the following year, we learned that the impact of eddies may be influenced by their origin, age, and period of the year. Finally, the metabolites with the strongest responses to eddy effects were unknown, and a relatively smaller proportion of known metabolites in the anticyclone demonstrates the utility of untargeted metabolomics in exploring environmental variation and the large amount of information that can still be gleaned from such datasets.

@@

```{r targ barplot relative and absolute}
targ_barplot_df <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Trimethylamine N-oxide", "TMAO", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Hydroxyisoleucine", "HO-Ile", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Glycine betaine", "GBT", compound_name)) %>%
  mutate(compound_name=str_remove(compound_name, "^L-")) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 9), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  group_by(depth, station, compound_name, tripl) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("15 m", "DCM", "175 m"))) %>%
  mutate(station=factor(station, levels=as.character(4:14)))
base_tripl_barplot <- targ_barplot_df %>%
  ggplot() +
  aes(x=tripl, y=nM, fill=compound_name) +
  geom_hline(yintercept = 0) +
  facet_grid(depth~station, scales="free_y", switch = "x") +
  scale_fill_manual(values = c(splividis(9), "grey50")) +
  labs(x="Station", y="Metabolite concentration (nM)", fill=NULL) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background.x = element_blank(), 
        panel.border = element_blank())
abs_targ_bar_gp <- base_tripl_barplot +
  geom_col(width=1) +
  ggtitle("A) Absolute concentrations of most abundant metabolites") +
  scale_y_continuous(expand = expansion()) +
  theme(legend.position="none", strip.text.x = element_blank(), axis.title.x = element_blank())
rel_targ_bar_gp <- base_tripl_barplot +
  geom_col(width=1, position = "fill") +
  scale_y_continuous(labels = scales::label_percent(), expand = expansion(0.1)) +
  theme(legend.position = "bottom") +
  guides(fill=guide_legend(nrow=2, byrow = TRUE)) +
  ggtitle("B) Relative composition of the metabolome")
comb_targ_rel_gp <- egg::ggarrange(abs_targ_bar_gp, rel_targ_bar_gp, ncol=1, draw=FALSE)

>>>>

This shift in community structure and productivity should result in a corresponding shift in the composition of the particulate organic matter produced. Eukaryotic organisms have distinct chemical fingerprints from those of cyanobacteria and heterotrophic bacteria [@Heal2021; @Durham2022] and the relative contribution of these different taxa to total biomass should reflect their chemical composition. However, recent work from @Harke2021 showed that overall community function may be robust across eddy types at the surface while @Gleich2024 detected significant differences in protistan community composition and metabolic potential down to 250 meters, with heterotrophy-associated transcripts found at higher abundances in the cyclone. These results raise questions about the effect of mesoscale features on particulate organic matter composition.

In this work, we directly tested for differences in the chemical composition of particulate matter due to changes in eddy state in the NPSG. We sampled from cyclonic and anticyclonic eddies spatially near each other and used targeted and untargeted mass spectrometry to explore changes in the composition of small, polar molecules. We expected to find that overall metabolite abundance would reflect shifts in biomass over multiple depths. We also expected to see enrichment in the deep chlorophyll maximum of cyclonic features for those metabolites especially abundant in eukaryotic organisms. Finally, we predicted that these patterns would be robust across years and sampling regimes for a reliable way to link these ocean features with the chemical composition of organic matter in the open ocean.

@@

Metabolomics samples from all three sample sets were extracted via a modified Bligh & Dyer [@Boysen2018], with the aqueous fraction run on a HILIC column paired with a Q-Exactive HF mass spectrometer. Metabolites for which an authentic standard was available and run alongside were integrated in Skyline (HEE and MESO-SCOPE eddy centers) and MSDIAL (MESO-SCOPE eddy transect). Untargeted analysis was performed using the R package XCMS with each feature manually inspected for quality assurance. [Wei: I think this paragraph belongs to Methods section. Will: Maybe just shorten it extensively? Ask Anitra for advice since Rachel wanted this here.]

@@

This shift in metabolite abundance was largely driven by a corresponding shift in biomass. Particulate carbon and summed metabolite concentration were tightly correlated across all samples and depths ($\beta = 24.9\pm1.82$ (SE), Pearson's $r$ = .813, p-value < 0.001, Supplemental figure 2). This overall trend was largely driven by higher values at 15 meters and the DCM than at depth, but when analyzed at each depth individually the general correspondence held (Supplemental figure 2). The DCM and 175 meter samples had the stronger trends ($\beta_{DCM} = 22.6\pm5.53$, $r= .593$, p-value < 0.001; $\beta_{175m} = 23.9\pm7.36$, $r= .503$, p-value = 0.003) while 15 meter samples showed no significant relationship ($\beta_{15m} = 3.4\pm12.9$, $r= .049$, p-value = 0.792). This meant that metabolites contributed a fixed fraction of carbon to the particulate pool with our 53 quantified metabolites representing between 1.5% and 5% of the carbon in the system (Figure 2) [Benedetto: I find this very informative. It would be great to get a ballpark sense of the percentages including the unknown metabolites, but I understand that it might be too uncertain considering the required assumptions on molecular formulas and concentration. I would like to know your wild guess, though]. Particulate nitrogen was tightly correlated with particulate carbon ($r=.933$) but the fraction of particulate nitrogen represented by the quantified metabolites was slightly higher across the transect with contributions typically between 2% and 6%.

@@

Given the strength of the metabolomic response to SLA at the DCM , we additionally analyzed samples taken with high-resolution depth around the DCM at the center of both eddy poles. Here, we found that the ~40 meter depth range and the transition between the eddies explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001, Figure 4). Additionally, both sample depth and SLA were correlated with the first principal component of the metabolite matrix ($r_{depth}=.766$, $r_{SLA}=.684$, fraction of variance explained by PC1 = 30.8%). The cyclonic samples in particular show a much larger spread than the anticyclonic ones, indicating larger sample variability in the cyclone DCM relative to the anticyclone. A depth gradient is visible along NMDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom (Figure 4).

@@

fk_targ %>%
  filter(compound_name%in%c("Homarine", "Trigonelline", "Dimethylsulfonioacetate", "Dimethylsulfoniopropionate", "Taurine", "Isethionic acid", "Arsenobetaine")) %>%
  left_join(filled_file_metadata) %>%
  group_by(filename) %>%
  mutate(nM=nM/sum(nM)) %>%
  ungroup() %>%
  filter(depth=="DCM") %>%
  nest(data=-compound_name) %>%
  mutate(aovout=map(data, ~broom::tidy(aov(.x$nM~.x$sla_class)))) %>%
  unnest(aovout) %>% filter(term==".x$sla_class") %>%
  mutate(p.value=p.adjust(p.value, method="fdr"))
  ggplot(aes(x=sla_class, y=nM)) +
  # geom_boxplot() +
  geom_point() +
  facet_wrap(~compound_name, scales="free_y")

@@

The eddy effects, on the other hand, were much less strong during this cruise. Of the six known metabolites with the strong enrichment in the cyclone at the DCM in the MESO-SCOPE cruise, only isethionic acid was significantly different in the HEE dataset (isethionic acid t-test p-value$_{FDR}$ = 0.043, other five p-values > 0.25). This difference in isethionic acid concentration was only found after normalizing each sample to the sum of all metabolites in the sample to control for biomass. Arsenobetaine was again found to strongly enriched in the anticyclone at the DCM both when normalizing to biomass and when not, just as in MESO-SCOPE. Surprisingly, the 173.0921 *m/z* mass feature was also slightly enriched but this time in the *cyclone* DCM with peak areas approximately 1.4 times the values in the anticyclone (t-test p-value = 0.067, Mann-Whitney p-value = 0.045). The 170.1176 *m/z* mass feature was not detected in the HEE samples.

@@

The Hawaiian Eddy Experiment (HEE) data upset many of the expectations [Wei: consider calculating the effect size, as mentioned above. Will: Not sure what Wei means by this. Should I do effect size calculations here? Or just above when we didn't see an effect?] we developed during the MESO-SCOPE cruise in the previous year. Most surprising was the large difference between the eddies detected at the *surface*, while the DCM samples were functionally indistinguishable. This contrasts directly with the results from MESO-SCOPE and the results in @Gleich2024, who found eddy-driven shifts in protistan community composition to be larger at depth than at the surface. However, @Dugenne2023 found differences in nitrogen fixation rate and nitrogen fixer composition varying widely throughout the upper water column. The large surface differences were especially surprising given the large inter-replicate differences between the HEE surface samples in this study, while the DCM samples tended to be much more consistent.

@@

The strong SLA signal detected among the 175 meter samples was partially surprising to us given our expectation about the strongest effect at the DCM where eddy effects lift large concentrations of nutrients above the 1% light level. Instead, we found more compounds overall to be significantly different across the eddy transect in the 175 meter samples than we did at the DCM (α=0.05, 43 compounds at 175 meters vs 22 at the DCM). This was likely driven by the enhancement of organic matter and heterotrophic picoplankton below the DCM as seen in @Barone2019, with the enrichment of nucleobases in the 175 meter samples additionally hint at increases in bacterial biomass given that nucleobases tend to be highly abundant in bacterial cultures [@Heal2021]. The abundance of a mass feature putatively identified as acetylcholine in the 175 meter cyclone samples followed the same general trend as arsenobetaine, both enriched at depth and in the anticyclone, and may be another compound that results from heterotrophic degradation of phytoplankton metabolites [@Durham2022]. These results raise important questions about the depth at which the community is isolated from SLA effects, with additional data from @Barone2022 and @Gleich2024 indicating that even at 250 meters eddy effects are still discernable.

@@

This study reports how changes in the biogeochemistry and community composition due to mesoscale eddies affect the metabolome and thereby the composition of particulate organic matter. A transect across adjacent mesoscale eddies of opposite polarity showed that many metabolites track closely with metrics of overall biomass, with eddy effects stronger at 175 meters than at 15 meters or the deep chlorophyll maximum (DCM). High-resolution depth sampling of the DCM at the center of each eddy elucidated several known and unknown biomarkers likely corresponding to the previously documented increase in eukaryotic phytoplankton in the cyclone. Metabolite clusters aligned well with expected trends in abundance with depth and between the two eddies, with convergence below the DCM towards a general deep-sea metabolite signal for many metabolites. By contrasting these effects with a follow-up analysis in the centers of a separate pair of mesoscale eddies of opposite polarity the following year, we learned that the impact of eddies may be influenced by their origin, age, and period of the year. Finally, the metabolites with the strongest responses to eddy effects were unknown, and a relatively smaller proportion of known metabolites in the anticyclone demonstrates the utility of untargeted metabolomics in exploring environmental variation and the large amount of information that is missed by only investigating known compounds.

@@

```{r targ barplot relative and absolute}
targ_barplot_df <- targ_peaks %>%
  left_join(filled_file_metadata, by="filename") %>%
  filter(ms_run=="MT") %>%
  filter(samp_type=="Smp") %>%
  group_by(compound_name) %>%
  mutate(mean_cmpd_nm=mean(nM)) %>%
  ungroup() %>%
  arrange(desc(mean_cmpd_nm)) %>%
  mutate(compound_name=ifelse(compound_name=="Trimethylamine N-oxide", "TMAO", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Hydroxyisoleucine", "HO-Ile", compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Glycine betaine", "GBT", compound_name)) %>%
  mutate(compound_name=str_remove(compound_name, "^L-")) %>%
  mutate(compound_name=ifelse(
    compound_name%in%head(unique(compound_name), 9), 
    compound_name, "Other")) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  group_by(depth, station, compound_name, tripl) %>%
  summarise(nM=sum(nM), .groups = "keep") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("15 m", "DCM", "175 m"))) %>%
  mutate(station=factor(station, levels=as.character(4:14)))
base_tripl_barplot <- targ_barplot_df %>%
  ggplot() +
  aes(x=tripl, y=nM, fill=compound_name) +
  geom_hline(yintercept = 0) +
  facet_grid(depth~station, scales="free_y", switch = "x") +
  scale_fill_manual(values = c(splividis(9), "grey50")) +
  labs(x="Station", y="Metabolite concentration (nM)", fill=NULL) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background.x = element_blank(), 
        panel.border = element_blank())
abs_targ_bar_gp <- base_tripl_barplot +
  geom_col(width=1) +
  ggtitle("A) Absolute concentrations of most abundant metabolites") +
  scale_y_continuous(expand = expansion()) +
  theme(legend.position="none", strip.text.x = element_blank(), axis.title.x = element_blank())
rel_targ_bar_gp <- base_tripl_barplot +
  geom_col(width=1, position = "fill") +
  scale_y_continuous(labels = scales::label_percent(), expand = expansion(0.1)) +
  theme(legend.position = "bottom") +
  guides(fill=guide_legend(nrow=2, byrow = TRUE)) +
  labs(y=NULL) +
  ggtitle("B) Relative composition of the metabolome")
comb_targ_rel_gp <- egg::ggarrange(abs_targ_bar_gp, rel_targ_bar_gp, ncol=1, draw=FALSE)
